\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-03-06 paper/main.tex
finish abstract

Here we introduce a framework that allows individuals to estimate their fitness
based on a minimum of social information. We explore the difference between our
framework and the classical framework using computer simulations. The
simulations show that in the prisoner's dilemma, individuals with limited memory
tend to adopt less generous strategies and they achieve less cooperation than in
the classical scenario.
\end{abstract}

@@

Evolutionary game theory~\cite{smith1982evolution, hofbauer1998evolutionary,
nowak:Nature:2004, hauert2005game} describes the evolutionary dynamics of
populations consisting of different types of interacting individuals. In the
last two decades the interest of the field has shifted to the analysis of
stochastic finite population dynamics in preference to traditional approaches
such as the replicator dynamics~\cite{hofbauer:JTB:1979, taylor:MATHBIO:1978}.
This is especially true in the topic of cooperation~\cite{hilbe:PNAS:2013,
hilbe:Nature:2018,glynatsi:SCR:2020}. In stochastic evolutionary dynamics
disadvantageous mutants have a small yet non-zero probability to reach fixation
and this can lead to fundamental changes in the results. As shown
in~\cite{nowak:Nature:2004}, using such dynamics a single cooperative strategy
can invade a population of defectors without any special modifications in
comparison to traditional approaches that one has to consider spatial
structure~\cite{nowak:Nature:1992} or when payoffs are subject to aggregate
shocks~\cite{fudenberg:JET:1992}.

@@

Similar to all other theoretical models, finite stochastic processes rest on a
set of assumptions. Namely, in the game phase it is assumed that players use
strategies with finite memory~\cite{Nowak1992tit, Baek2016}. For example, in a
repeated game between two individuals the actions of the players in each turn
are often determined by the action of the co-players in the previous turn. This
assumption is common as it allows for an explicit calculation of the players'
payoffs~\cite{sigmund2010calculus}. In addition, it is
assumed that individuals play many times and with all other players before
reproduction takes place. So that in the updating phase the updating payoffs, and
subsequently the fitness, of individuals is given by the average payoffs of an
infinitely repeated game against the mean distribution of types in the
population.

@@

We first consider two extreme scenarios, the classical scenario where an individual
has a perfect memory of all their interactions and the alternative scenario of
limited memory where individuals update their strategies only based on the very
last payoff they obtained. We demonstrate the effects of limited payoff
memory using a pairwise comparison process and the well studied game of the
prisoner's dilemma. We observe that individuals with limited payoff memory tend to
adopt less generous strategies and they achieve less cooperation when
interacting in a prisoner's dilemma. We obtain similar results when we consider
that individuals update their strategies based on more information. More
specifically, up to the last two payoffs they obtained when interacting with up
to two different members of the population.

@@

A pairwise comparison process starts with assigning all
individuals of the population the same strategy. Thenceforth each elementary time step of
the process consists of the mutation phase, the game phase and the update phase.
In the game phase individuals are matched in pairs and that they participate in
a repeated 2 person donation game; a special case of the prisoner's dilemma. In
the donation game there are two actions: cooperation (\(C\)) and defection
(\(D\)). By cooperating a player provides a benefit \(b\) to the other player at
their cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn
are given by,

@@

In repeated games there are infinite many strategies, however, similar to the
literature we will assume that individuals use reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently. The play between a
pair of reactive strategies $s_1 = (y_1, p_1, q_1)$ and $s_2 = (y_2, p_2, q_2)$
can be model as a Markov process with the transition matrix \(M\),

@@

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the game stage, which happens with
probability $\frac{1}{(N\!-\!1)}$.

@@

In order to account for the effect of the updating payoffs we simulate the
evolutionary process and record which strategies the players adopt over time
based on perfect and limited payoff memory. We performed two independent runs
for each approach varying the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$ (since the discount
factor~$\delta$ is comparably high we do not report the opening move \(y\)). The
upper panel of Figure~\ref{fig:expected_and_stochastic_for_donation} corresponds
to the standard scenario considered in the literature, and the bottom panel
shows the scenario considered herein, in which players update their strategies
based on their last round's payoff.

@@

We further explore the effects of the cooperation benefit and the strength of
selection on the evolved generosity \(q\) and on the cooperation rate
(Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}).
Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{A} suggests that
perfect memory always yields a higher cooperation rate. In this case the
cooperation rate increases as the benefit increases. In comparison for limited
payoff memory the cooperation rate remains unchanged at approximately 50\% once
\(b=5\).  From Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{B}
we observe that for weak selection, \(\beta < 1\), the two methods yield similar
results, however, as \(\beta\) increases there is variation in the evolving
populations. In the case of expected payoffs the resident populations become
more cooperative, whereas in the case of limited payoff memory, the resident
populations become more defective.

@@

Up to this point we have discussed only two edge cases. The framework of limited
payoff memory can be easily extended to allow individuals to observe more
rounds, to interact with more members or both (see Supplementary Information
Sections 4-6). To further understand the effect of limited payoff memory we
consider the cases of updating based on the last round with two members of the
population, the last two rounds with another member of the population, and the
last two rounds with two members of the population. Similar to the last round
payoff. We perform numerical simulations for each of the fitness approaches. In
Figure~\ref{fig:cooperation_rate_all_updating_payoffs} we report the cooperation
rates for low and high values of benefit.

@@

Here, we have explored how robust cooperation is as models deviate from the
perfect memory assumption. We considered a special case of the prisoner's
dilemma, the donation game. The prisoner's dilemma is one of the most applied
types of social games when we discuss the evolution of cooperation. We showed
that when the last round payoff is used instead of the expected payoffs,
cooperation struggles to evolve  when only a minimum of social information is
used at the updating stage. The resident strategies were both less cooperative
and less generous. This effect was only intensified we increase the benefit and
the strength of selection independently. The results showed that as each
parameter was being increased the difference between the cooperation rates were
widening. This indicates that cooperative players benefit from being able to
interact with everyone in the population.

>>>>

Here we introduce a framework that allows individuals to estimate their fitness
based on a minimum of social information. We explore the difference between our
framework and the classical framework using the iterated prisoner's dilemma. We
show that individuals relying on a single set of information tend to adopt less
generous strategies and they achieve less cooperation than in the classical
scenario. Although strategies in the latter scenario become more cooperative
with increasing benefits, those in the former remain unchanged. We further
explore scenarios where individuals are granted slightly more information, and
our observations indicate that these individuals can attain higher levels of
cooperation.

\end{abstract}

@@

Evolutionary game theory~\cite{smith1982evolution, hofbauer1998evolutionary,
nowak:Nature:2004, hauert2005game} describes the evolutionary dynamics of
populations consisting of different types of interacting individuals. In the
last two decades the interest of the field has shifted to the analysis of
stochastic finite population dynamics in preference to traditional approaches
such as the replicator dynamics~\cite{hofbauer:JTB:1979, taylor:MATHBIO:1978}.
This is especially true in the topic of cooperation~\cite{hilbe:PNAS:2013,
hilbe:Nature:2018,glynatsi:SCR:2020, ohtsuki:Nature:2006, nowak:Science:2006,
murase:ScientificReports:2020}. In stochastic evolutionary dynamics
disadvantageous mutants have a small yet non-zero probability to reach fixation
and this can lead to fundamental changes in the results. As shown
in~\cite{nowak:Nature:2004}, using such dynamics a single cooperative strategy
can invade a population of defectors without any special modifications in
comparison to traditional approaches that one has to consider spatial
structure~\cite{nowak:Nature:1992} or when payoffs are subject to aggregate
shocks~\cite{fudenberg:JET:1992}.

@@

Similar to all other theoretical models, finite stochastic processes rest on a
set of assumptions. Namely, in the game phase it is assumed that players use
strategies with finite memory~\cite{Nowak1992tit, Baek2016}. For example, in a
repeated game between two individuals the actions of the players in each turn
are often determined by the action of the co-players in the previous turn~\cite{Nowak1992tit}. This
assumption is common as it allows for an explicit calculation of the players'
payoffs~\cite{sigmund2010calculus}. In addition, it is
assumed that individuals play many times and with all other players before
reproduction takes place. So that in the updating phase the updating payoffs,
or fitness, of individuals is given by the average payoffs of an
infinitely repeated game against every member of the population.

@@

We first consider two extreme scenarios, the classical scenario where an individual
has a perfect memory of all their interactions and the alternative scenario of
limited memory where individuals update their strategies only based on the very
last payoff they obtained. We demonstrate the effects of limited payoff
memory using a pairwise comparison process and the well studied game of the
prisoner's dilemma~\cite{glynatsi:HSSC:2021}. We observe that individuals with
limited payoff memory tend to adopt less generous strategies. We obtain similar
results when we consider that individuals update their strategies based on more
information. More specifically, up to the last two payoffs they obtained when
interacting with up to two different members of the population.

@@

A pairwise comparison process~\cite{traulsen2007pairwise} starts with assigning all
individuals of the population the same strategy. Thenceforth each elementary time step of
the process consists of the mutation phase, the game phase and the update phase.
In the game phase individuals are matched in pairs and that they participate in
a repeated 2 person donation game; a special case of the prisoner's dilemma. In
the donation game there are two actions: cooperation (\(C\)) and defection
(\(D\)). By cooperating a player provides a benefit \(b\) to the other player at
their cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn
are given by,

@@

In repeated games there are infinite many strategies, however, similar to the
literature we will assume that individuals use reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently. The play between a
pair of reactive strategies $s_1 = (y_1, p_1, q_1)$ and $s_2 = (y_2, p_2, q_2)$
can be model as a Markov process with the transition matrix \(M\)~\cite{nowak:APC:1989},

@@

@@

In order to account for the effect of the updating payoffs we simulate the
evolutionary process and record which strategies the players adopt over time
based on perfect and limited payoff memory. We performed two independent runs
for each approach varying the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$ (since the discount
factor~$\delta$ is comparably high we do not report the opening move \(y\)).

@@

We further explore the effects of the benefit and the strength of
selection on the generosity \(q\) and on the cooperation rate
(Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}).
Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{A} suggests that
perfect memory always yields a higher cooperation rate. In this case the
cooperation rate increases as the benefit increases. In comparison for limited
payoff memory the cooperation rate remains unchanged at approximately 50\% once
\(b=5\).  From Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{B}
we observe that for weak selection, \(\beta < 1\), the two methods yield similar
results, however, as \(\beta\) increases there is variation in the evolving
populations. In the case of expected payoffs the resident populations become
more cooperative, whereas in the case of limited payoff memory, the resident
populations become more defective.

@@

The framework of limited
payoff memory can be easily extended to allow individuals to observe more
rounds, to interact with more members or both (see Supplementary Information
Sections 4-6). To further understand the effect of limited payoff memory we
consider the cases of updating based on the last round with two members of the
population, the last two rounds with another member of the population, and the
last two rounds with two members of the population. We perform numerical
simulations for each of the fitness methods. In
Figure~\ref{fig:cooperation_rate_all_updating_payoffs} we report the cooperation
rates for low and high values of benefit.

@@

Here, we have explored how robust cooperation is as models deviate from the
perfect memory assumption. We considered a special case of the prisoner's
dilemma, the donation game. We showed
that when the last round payoff is used instead of the expected payoffs,
cooperation struggles to evolve  when only a minimum of social information is
used at the updating stage. The resident strategies were both less cooperative
and less generous. This effect was only intensified we increase the benefit and
the strength of selection independently. The results showed that as each
parameter was being increased the difference between the cooperation rates were
widening. This indicates that cooperative players benefit from being able to
interact with everyone in the population.
